---
title: "isocat"
author: "Caitlin J. Campbell"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    keep_md: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(kableExtra)
```

**Contact Information:** Caitlin Campbell ([caitjcampbell@gmail.com](caitjcampbell@gmail.com))


# Overview

The `isocat` package adds multiple tools for creating and quantitatively analyzing and summarizing probability-of-origin surfaces generated from stable isotope analyses of animal tissue. This vignette will walk users through a brief example for each function included in this vignette.

```{r load isocat}
library(isocat)
```

# Loading example data

`isocat` has example isoscape data included for a small extent of North America.

Example isoscape:

```{r load isoscape data}
data(isoscape)
myiso <- rasterFromXYZ(isoscape)
raster::plot(myiso)
```

Example isoscape standard deviation surface:

```{r load isoscape data 2}
myiso_sd <- rasterFromXYZ(isoscape_sd)
```

# Creating a probability-of-origin surface

To create a probability-of-origin map, we first import or generate a dataframe containing:

1) Individual IDs

2) Individual-level isotope data, transformed with a transfer function to reflect isoscape values.[^1]

3) Standard deviation of isotope standard measurements, which are associated with machine accuracy.

[^1]: [TODO]  

```{r example dataframe}
df <- data.frame(ID = LETTERS[1:3], dD = c(-100, -80, -50), SD_indv = rep(5, 3))
kable(df)
```

The contents of these columns are passed to the function `isotopeAssignmentModel` as vectors, along with the object names of isoscape and isoscape-SD objects. If parallel processing is specified, the function creates and deploys clusters using the `doParallel` package. The output is a rasterStack with layers named corresponding to the individual IDs.

```{r prob of orgin surface}
assignmentModels <- isotopeAssignmentModel(
  ID = df$ID, 
  dD = df$dD, 
  SD_indv = df$SD_indv, 
  precip_raster = myiso, 
  precip_SD_raster = myiso_sd, 
  nClusters = FALSE
  )
raster::plot(assignmentModels)
```

# Comparing surfaces

To compare probability-of-origin surfaces, we apply Schoener's D metric. To simply compare two surfaces, we can apply `isocat`'s `schoenersD` function, which determine's Schoener's D-metric of similarity between two surfaces. The D-value varies between 0 (completely dissimilar surfaces) and 1 (identical surfaces).

```{r schoenersD}
# Calculate Schoener's D-metric of spatial similarity between two of the example probability surfaces.
schoenersD(assignmentModels[[1]], assignmentModels[[2]])
```

To compare multiple surfaces to one another, `isocat` includes a `simmatrixMaker` function to create a similarity matrix of the surfaces. The output is a symmetric matrix with row and column names corresponding to the layernames of the surfaces to be compared. The `nClusters` specification, as in the `isotopeAssignmentModel` function, generates a number of parallel processing clusters equal to the numeric value specified. If `csvSavePath` is included, a .csv file will also be written to the path specified. For large rasterStacks, this function can be quite processing-intensive and take some time.

```{r simmatrix}
simmatrixMaker(
  assignmentModels,
  nClusters = FALSE,
  csvSavePath = FALSE
)
```
# Clustering

To cluster individuals by similar origins, 

# Post-processing surfaces

