---
title: "Summarizing origin environments from probability-of-origin surfaces"
author: "C.J. Campbell"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    keep_md: true
vignette: >
  %\VignetteIndexEntry{Summarizing origin environments from probability-of-origin surfaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(isocat)
library(terra)
library(ggplot2)
library(kableExtra)
```

**Contact Information:** Caitlin J. Campbell (<caitjcampbell@gmail.com>)

# Overview

Once you have a probability-of-origin surface for an individual, a common next
question is *what were the plausible origin conditions?* — the environment
(elevation, climate, an isoscape value) an individual likely came from, and how
far it likely travelled from where it was sampled.

`isocat` answers this by drawing a probability-weighted cloud of candidate origin
points from each surface, attaching environmental values to those points, and
reducing the point cloud to per-individual summaries. The workflow is four small
functions:

1. `sampleOriginPoints()` — draw points from a surface, weighted by probability of
   origin, optionally relative to a sampling location.
2. `extractEnvPoints()` — attach environmental covariates at each point.
3. `summarizeDistribution()` — reduce the retained point cloud to per-individual
   summaries, optionally under a movement prior.
4. `plotOriginDistribution()` — visualize the per-individual distributions.

Because the points are drawn *with probability proportional to the surface*, the
probability weighting is already baked into the sample, and every downstream
summary is a simple (optionally movement-weighted) average over points.

```{r load}
library(isocat)
library(terra)
```

# Example data and a simulated sampling site

We build a small set of probability-of-origin surfaces from the bundled isoscape,
following the approach in the main vignette.

```{r surfaces, fig.width = 6, fig.height = 3}
myiso    <- rast(isoscape,    type = "xyz")
myiso_sd <- rast(isoscape_sd, type = "xyz")

df <- data.frame(
  ID           = LETTERS[1:5],
  isotopeValue = c(-120, -100, -80, -60, -45),
  SD_indv      = rep(5, 5)
)

surfaces <- isotopeAssignmentModel(
  ID               = df$ID,
  isotopeValue     = df$isotopeValue,
  SD_indv          = df$SD_indv,
  precip_raster    = myiso,
  precip_SD_raster = myiso_sd,
  nClusters        = FALSE
)
plot(surfaces)
```

For a worked example we imagine every individual was captured at a single
sampling site placed at the **center of the study extent**. In a real analysis this
would be each individual's true capture coordinates.

```{r site}
ev   <- as.vector(terra::ext(surfaces))            # xmin, xmax, ymin, ymax
site <- c(x = mean(ev[1:2]), y = mean(ev[3:4]))
site
```

We also assemble an environmental stack aligned to the surfaces. Here we use the
isoscape itself as a real covariate (`d2H`) and add a simple synthetic gradient
(`gradient`, increasing with latitude) purely to illustrate multi-layer
extraction.

```{r env}
env <- c(myiso, terra::init(myiso, "y"))
names(env) <- c("d2H", "gradient")
```

# Example 1: Sample probability-weighted origin points

`sampleOriginPoints()` draws `n` points from each layer, sampling cells with
probability proportional to their value. Passing `xy` (our sampling site) adds each
point's easting, northing, and Euclidean distance from the site, in the surface's
map units. Set a seed for reproducibility.

```{r sample}
set.seed(42)
pts <- sampleOriginPoints(surfaces, n = 1000, xy = site)
head(pts)
```

The result is one row per point. Plotted over a single individual's surface, the
cloud is denser where the probability of origin is higher:

```{r cloud, fig.width = 6, fig.height = 4}
plot(surfaces[["A"]], main = "Individual A: surface and sampled origin points")
points(pts[pts$ID == "A", c("x", "y")], pch = 16, cex = 0.2,
       col = adjustcolor("black", alpha.f = 0.2))
points(site["x"], site["y"], pch = 4, cex = 1.5, lwd = 2, col = "red")
```

# Example 2: Extract environmental values at the points

`extractEnvPoints()` reads the environmental stack at each point's coordinates and
appends one column per layer. The full point-by-point distribution is retained.

```{r extract}
pts <- extractEnvPoints(pts, env)
head(pts)
```

# Example 3: Summarize the origin distribution per individual

`summarizeDistribution()` reduces the points to per-individual, per-variable
summaries: a mean, standard deviation, and quantiles. Alongside the environmental
variables we can summarize `dist` directly to describe probable distance travelled.

```{r summarize}
summ <- summarizeDistribution(pts, vars = c("d2H", "gradient", "dist"))
kableExtra::kable(summ, digits = 2)
```

The `d2H` summary recovers each individual's plausible origin isoscape value, and
the `dist` summary describes how far, in map units, its plausible origins fall from
the sampling site.

# Example 4: Apply a movement prior

Distant origins are often less plausible than nearby ones. A finite movement scale
`L` down-weights points far from the sampling site with a half-normal
`movementKernel()`; `L = Inf` (the default) leaves the summary unweighted. The
kernel is the *only* explicit weight, since probability weighting is already in the
sample.

```{r kernel, fig.width = 5, fig.height = 3}
d <- seq(0, 15, length.out = 100)
plot(d, movementKernel(d, L = 5), type = "l", lwd = 2,
     xlab = "Distance from site (map units)", ylab = "Weight",
     main = "Half-normal movement kernel, L = 5")
```

Compare the unweighted origin `d2H` to the movement-weighted estimate. The movement
prior pulls each individual's estimate toward the environment nearer the site.

```{r weighted}
unweighted <- summarizeDistribution(pts, vars = "d2H", L = Inf)
weighted   <- summarizeDistribution(pts, vars = "d2H", L = 5)

comparison <- data.frame(
  ID            = unweighted$ID,
  d2H_unweighted = round(unweighted$mean, 1),
  d2H_L5         = round(weighted$mean, 1)
)
kableExtra::kable(comparison)
```

# Example 5: Visualize per-individual distributions

`plotOriginDistribution()` shows the spread of a variable across each individual's
sampled origins as horizontal violins — a ridgeline-style view of the point cloud.

```{r plot_dist, fig.width = 6, fig.height = 4}
plotOriginDistribution(pts, var = "d2H") +
  labs(x = "Origin d2H", y = "Individual",
       title = "Distribution of plausible origin d2H per individual")
```

The same call works for distance travelled, or any extracted covariate:

```{r plot_dist2, fig.width = 6, fig.height = 4}
plotOriginDistribution(pts, var = "dist") +
  labs(x = "Distance from sampling site (map units)", y = "Individual")
```

# Notes

- **Coordinate system.** Distances are Euclidean in the surface's map units, so use
  a projected (ideally equal-area) surface for meaningful distances and movement
  scales. The bundled isoscape is in geographic (lon/lat) degrees, so distances
  here are illustrative degrees.
- **Environmental alignment.** `extractEnvPoints()` reads values at point
  coordinates, so the environmental stack need not share the surface's grid — only
  its coordinate system.
- **Reproducibility.** `sampleOriginPoints()` uses `sample.int()`; call `set.seed()`
  beforehand to reproduce a point cloud.
