---
title: "Summarizing origin environments from probability-of-origin surfaces"
author: "C.J. Campbell"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    keep_md: true
vignette: >
  %\VignetteIndexEntry{Summarizing origin environments from probability-of-origin surfaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
library(isocat)
library(terra)
library(ggplot2)
library(kableExtra)
```

**Contact Information:** Caitlin J. Campbell (<caitjcampbell@gmail.com>)

# Overview

Once you have a probability-of-origin surface for an individual, a common next
question is *what were the plausible origin conditions?* — the environment
(elevation, climate, an isoscape value) an individual likely came from, and how
far it likely travelled from where it was sampled.

`isocat` answers this by folding the capture location into the surface, drawing a
probability-weighted cloud of candidate origin points, attaching environmental values
to those points, and reducing the cloud to per-individual summaries. The workflow is:

1. `competitiveSurface()` — reweight the surface by the capture location so nearby
   origins are favoured over distant ones *only when a closer origin is equally or more
   isotopically suitable* (the recommended, parameter-free way to bring in geography).
2. `sampleOriginPoints()` — draw points from a surface, weighted by probability of
   origin, recording each point's position relative to the capture site.
3. `extractEnvPoints()` — attach environmental covariates at each point.
4. `summarizeDistribution()` — reduce the retained point cloud to per-individual
   summaries.
5. `plotOriginDistribution()` — visualize the per-individual distributions.

Because the points are drawn *with probability proportional to the surface*, the
probability weighting is already baked into the sample, and every downstream summary is a
simple average over points.

There are two ways to bring the capture location into the summary. The recommended default
is **competitive distance reweighting** (`competitiveSurface()`), which needs no movement
scale and never discounts the most-suitable origin, wherever it lies. A parametric
**half-normal movement prior** (`movementKernel()`) is available as an alternative and is
shown at the end.

```{r load}
library(isocat)
library(terra)
```

# Example data and a simulated sampling site

We build a small set of probability-of-origin surfaces from the bundled isoscape,
following the approach in the main vignette.

Probability-of-origin surfaces should be in an **equal-area** projection so that
distances and areas derived from them are meaningful. The bundled isoscape already ships
on a square equal-area (Albers) grid in metres, with the CRS carried in a `"crs"`
attribute; we apply it when building the raster. (If you start from a lon/lat isoscape,
project it to an equal-area CRS first.)

```{r surfaces, fig.width = 7.5, fig.height = 5}
myiso    <- rast(isoscape,    type = "xyz", crs = attr(isoscape,    "crs"))
myiso_sd <- rast(isoscape_sd, type = "xyz", crs = attr(isoscape_sd, "crs"))

df <- data.frame(
  ID           = LETTERS[1:5],
  isotopeValue = c(-120, -100, -80, -60, -45),
  SD_indv      = rep(5, 5)
)

surfaces <- isotopeAssignmentModel(
  ID               = df$ID,
  isotopeValue     = df$isotopeValue,
  SD_indv          = df$SD_indv,
  precip_raster    = myiso,
  precip_SD_raster = myiso_sd,
  nClusters        = FALSE
)
plot(surfaces)
```

For a worked example we imagine every individual was captured at a single
sampling site placed at the **center of the study extent**. In a real analysis this
would be each individual's true capture coordinates.

```{r site}
ev   <- as.vector(terra::ext(surfaces))            # xmin, xmax, ymin, ymax
site <- c(x = mean(ev[1:2]), y = mean(ev[3:4]))
site
```

We also assemble an environmental stack aligned to the surfaces. Here we combine two
real surfaces: the isoscape itself (`d2H`) and a digital elevation model. `isocat`
bundles `example_dem`, elevation on the same equal-area grid as the isoscape (from NOAA's
public-domain ETOPO 2022 global relief model), so it stacks directly. A secondary
`SpatRaster` on a different grid would first be projected to match with
`project(dem, myiso)`.

```{r env}
dem <- rast(example_dem, type = "xyz", crs = attr(example_dem, "crs"))
env <- c(myiso, dem)
names(env) <- c("d2H", "elevation")
```

# Example 1: Competitive distance reweighting

Isotopic suitability alone treats an origin 2000 km away exactly like an equally-suitable
origin next to the capture site. `competitiveSurface()` folds in the capture location so
that a candidate origin is discounted **only when a closer, equally- or more-suitable
origin exists**. Formally, each cell is weighted by the isotope-weighted probability that
an equally- or more-suitable location lies at least as far from capture as that cell — a
within-suitability distance survival function in $[0, 1]$.

This has three useful properties:

- **Isotope-led.** The single most-suitable cell always keeps weight 1 and is never
  discounted, *wherever it is* — so a genuine long-distance origin survives.
- **Competitive.** A cell is penalized only by locations that are *both* more suitable
  *and* nearer; diffuse low-suitability tails dissolve because each such cell has a
  superior, nearer alternative.
- **Parameter-free.** It uses only the *ordering* of distances, so — unlike a movement
  kernel — there is no movement scale to choose, and the result is invariant to the
  absolute distance scale.

```{r reweight, fig.width = 7.5, fig.height = 5}
psurf <- competitiveSurface(surfaces, xy = site)
plot(psurf)
```

The effect is clearest side by side. The raw surface for individual C spreads probability
across every isotopically-suitable cell; the reweighted surface pulls that mass toward the
capture site (red X), retaining distant cells only where they are the most suitable
available.

```{r reweight_compare, fig.width = 8, fig.height = 4}
op <- par(mfrow = c(1, 2))
plot(surfaces[["C"]], main = "Raw isotopic surface (C)")
points(site["x"], site["y"], pch = 4, cex = 1.5, lwd = 2, col = "red")
plot(psurf[["C"]], main = "Competitively reweighted (C)")
points(site["x"], site["y"], pch = 4, cex = 1.5, lwd = 2, col = "red")
par(op)
```

The rest of the workflow runs on the reweighted surface `psurf`.

# Example 2: Sample probability-weighted origin points

`sampleOriginPoints()` draws `n` points from each layer, sampling cells with
probability proportional to their value. Passing `xy` (our sampling site) adds each
point's easting, northing, and Euclidean distance from the site in the surface's map
units — metres, since the surface is equal-area. We add kilometre versions for
readability. Set a seed for reproducibility.

```{r sample}
set.seed(42)
pts <- sampleOriginPoints(psurf, n = 1000, xy = site)
pts$easting_km  <- pts$easting  / 1000
pts$northing_km <- pts$northing / 1000
pts$dist_km     <- pts$dist     / 1000
head(pts)
```

The result is one row per point, across all five individuals. Taking individual C as an
example, we can confirm the draws represent its surface by placing them side by side on the
same spatial scale: the reweighted surface with the sampled points overlaid (left), and a
2-D kernel density of those same points (right). The smoothed cloud recovers the surface —
denser where the probability of origin is higher — with the sampling site marked by the
red X.

```{r cloud, fig.width = 9, fig.height = 4.6, eval = requireNamespace("gridExtra", quietly = TRUE)}
ptsC  <- pts[pts$ID == "C", ]
csurf <- as.data.frame(psurf[["C"]], xy = TRUE)          # reweighted surface as x/y/prob
names(csurf)[3] <- "prob"
siteC <- data.frame(x = site[["x"]], y = site[["y"]])

# Shared spatial scale + look for both panels.
xr <- range(csurf$x); yr <- range(csurf$y)
panel <- list(
  coord_equal(xlim = xr, ylim = yr, expand = FALSE),
  scale_fill_viridis_c(option = "viridis"),
  theme_minimal(base_size = 11),
  theme(axis.text = element_blank(), axis.title = element_blank(),
        axis.ticks = element_blank(), panel.grid = element_blank(),
        plot.subtitle = element_text(face = "bold"))
)
star <- geom_point(data = siteC, aes(x, y), shape = 4, size = 3, stroke = 1.2,
                   colour = "red", inherit.aes = FALSE)

p_surface <- ggplot(csurf, aes(x, y)) +
  geom_raster(aes(fill = prob)) +
  geom_point(data = ptsC, aes(x, y), colour = "white", alpha = 0.15, size = 0.35,
             inherit.aes = FALSE) +
  star + panel +
  labs(subtitle = "Reweighted surface + sampled points", fill = "P(origin)")

# Estimate the density over the full surface extent so the raster fills the panel.
kd  <- MASS::kde2d(ptsC$x, ptsC$y, n = 150, lims = c(xr, yr))
kdf <- expand.grid(x = kd$x, y = kd$y); kdf$density <- as.vector(kd$z)
p_density <- ggplot(kdf, aes(x, y, fill = density)) +
  geom_raster() +
  star + panel +
  labs(subtitle = "Kernel density of sampled points", fill = "Density")

gridExtra::grid.arrange(p_surface, p_density, nrow = 1)
```

# Example 3: Extract environmental values at the points

`extractEnvPoints()` reads the environmental stack at each point's coordinates and
appends one column per layer. Our stack holds two surfaces — the isoscape (`d2H`) and
the ETOPO elevation (`elevation`) — so a single call attaches both, and the DEM value
now rides along on every sampled origin. The full point-by-point distribution is
retained.

```{r extract}
pts <- extractEnvPoints(pts, env)
head(pts[, c("ID", "x", "y", "dist", "d2H", "elevation")])
```

# Example 4: Summarize the origin distribution per individual

`summarizeDistribution()` reduces the points to per-individual, per-variable
summaries: a mean, standard deviation, and quantiles. The summary is a plain average over
the points — no explicit weighting, because both the probability of origin and the
competitive distance reweighting are already contained in the sampled cloud.

```{r summarize}
summ <- summarizeDistribution(pts, vars = c("d2H", "elevation"))
kableExtra::kable(summ, digits = 2)
```

The `d2H` summary recovers each individual's plausible origin isoscape value, and the
`elevation` summary gives the probable elevation of origin (m).

The same call summarizes the *spatial geometry* of the plausible origins relative to the
sampling site. The `q2.5`–`q97.5` span is the range of each (in km): how far east/west
(`easting_km`) and north/south (`northing_km`) origins fall, and the spread of
straight-line distance travelled (`dist_km`).

```{r summarize_spatial}
spatial <- summarizeDistribution(pts, vars = c("easting_km", "northing_km", "dist_km"))
kableExtra::kable(spatial, digits = 1)
```

# Example 5: Visualize per-individual distributions

`plotOriginDistribution()` shows the spread of a variable across each individual's
sampled origins, one row per individual. The default `type = "halfeye"` draws a
[ggdist](https://mjskay.github.io/ggdist/) slab-and-interval: a density with a
median-and-quantile-interval beneath it. Because the sampled points are Monte-Carlo
draws from the probability surface, these distribution-of-draws idioms apply directly.

```{r plot_dist, fig.width = 6, fig.height = 4, eval = requireNamespace("ggdist", quietly = TRUE)}
plotOriginDistribution(pts, var = "d2H") +
  labs(x = "Origin d2H", y = "Individual",
       title = "Distribution of plausible origin d2H per individual")
```

The same call works for the elevation extracted from the DEM, or any other covariate.
Other `type`s include `"eye"`, `"dots"` (a quantile dotplot, also via ggdist), and
`"violin"` (a plain ggplot2 violin, requiring only ggplot2):

```{r plot_dist2, fig.width = 6, fig.height = 4, eval = requireNamespace("ggdist", quietly = TRUE)}
plotOriginDistribution(pts, var = "elevation", type = "dots") +
  labs(x = "Origin elevation (m)", y = "Individual")
```

`plotOriginDistribution()` shows one variable at a time. To view several together — here
the spatial geometry of each individual's origins — reshape the points to long form and
facet. This gives a per-individual view of easting, northing, and distance travelled; the
dashed line marks the sampling site (0), so easting/northing show which way, and how far,
plausible origins lie from it.

```{r plot_spatial, fig.width = 8, fig.height = 4, eval = requireNamespace("ggdist", quietly = TRUE)}
spatial_long <- data.frame(
  ID       = rep(pts$ID, 3),
  variable = factor(rep(c("Easting (km)", "Northing (km)", "Distance (km)"),
                        each = nrow(pts)),
                    levels = c("Easting (km)", "Northing (km)", "Distance (km)")),
  value    = c(pts$easting_km, pts$northing_km, pts$dist_km)
)
ggplot(spatial_long, aes(x = value, y = ID)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
  ggdist::stat_halfeye(fill = "steelblue", alpha = 0.8) +
  facet_wrap(~ variable, scales = "free_x") +
  labs(x = NULL, y = "Individual",
       title = "Per-individual distribution of origin geometry") +
  theme_bw()
```

# Alternative: a parametric movement prior

Competitive reweighting is parameter-free, but sometimes you have an independent estimate
of how far a species can move and want to impose it explicitly. `movementKernel()`
provides a half-normal movement prior, $\pi(d) = \exp(-0.5 (d / L)^2)$, at a movement
scale `L` you supply. It is applied as the *only* explicit weight in
`summarizeDistribution()` (via `L=`), on points drawn from the **raw** surface — so it is
an alternative to `competitiveSurface()`, not a complement.

```{r kernel, fig.width = 5, fig.height = 3}
d <- seq(0, 800, length.out = 100)
plot(d, movementKernel(d, L = 300), type = "l", lwd = 2,
     xlab = "Distance from site (km)", ylab = "Weight",
     main = "Half-normal movement kernel, L = 300 km")
```

Sample from the raw surface, then compare the unweighted origin `d2H` to the
movement-weighted estimate (here a 300 km scale, applied to `dist_km`). The movement prior
pulls each estimate toward the environment nearer the site.

```{r weighted}
set.seed(42)
pts_raw <- sampleOriginPoints(surfaces, n = 1000, xy = site)
pts_raw$dist_km <- pts_raw$dist / 1000
pts_raw <- extractEnvPoints(pts_raw, env)

unweighted <- summarizeDistribution(pts_raw, vars = "d2H", L = Inf)
weighted   <- summarizeDistribution(pts_raw, vars = "d2H", L = 300, dist = "dist_km")

comparison <- data.frame(
  ID             = unweighted$ID,
  d2H_unweighted = round(unweighted$mean, 1),
  d2H_L300km     = round(weighted$mean, 1)
)
kableExtra::kable(comparison)
```

The two approaches answer different questions. Competitive reweighting asks *is there a
better, nearer origin?* and needs no scale; the movement kernel asks *how far is this
species likely to have moved?* and requires you to commit to an `L`. Use the movement
kernel when you have a defensible movement scale; otherwise the parameter-free competitive
surface is the safer default.

# Notes

- **Coordinate system.** `easting`, `northing`, `dist`, and the competitive reweighting
  are all Euclidean in the surface's map units, so surfaces should be in an equal-area
  projection for these to be meaningful. The bundled isoscape already ships on a square
  equal-area (Albers) grid in metres, with its CRS carried in a `"crs"` attribute. If you
  start from a lon/lat isoscape, project it to equal-area first; both
  `competitiveSurface()` and `sampleOriginPoints()` warn if handed a lon/lat surface.
- **Environmental alignment.** `extractEnvPoints()` reads values at point
  coordinates, so the environmental stack need not share the surface's grid — but it
  must share its coordinate system (project secondary surfaces to match).
- **Reproducibility.** `sampleOriginPoints()` uses `sample.int()`; call `set.seed()`
  beforehand to reproduce a point cloud.
